Well-quasi-orderings on word languages

### 2024-08-15 src/infixes.tex
typos

In this section, we study the well-quasi-ordering of languages under the
\kl{infix relation}. As opposed to the \kl{prefix} and \kl{suffix} relations,
the \kl{infix relation} can yield to very complicated \kl{well-quasi-ordered}
languages. Formally, the upcoming \cref{infix-embedding:thm} shows that \emph{any} countable
quasi-ordering with finite initial segments can be embedded into the infix
relation of a language.

\AP Let us recall that an \intro{order embedding} from a quasi-ordered set $(X,
\preceq)$ into a quasi-ordered set $(Y, \preceq')$ is a function $f \colon X
\to Y$ such that for all $x, y \in X$, $x \preceq y$ if and only if $f(x)
\preceq' f(y)$. When such an embedding exists, we say that $X$ \reintro{embeds
into} $Y$. We say that a quasi-ordered set $(X, \preceq)$ is a \intro{partial
ordering} whenever the relation $\preceq$ is antisymmetric, that is $x \preceq
y$ and $y \preceq x$ implies $x = y$.

@@

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, it is clear that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} b^* c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon Y_n \to
    L_n$.

@@

\begin{theorem}[restate=infix-finite-automata:thm,label=infix-finite-automata:thm]
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

@@

@@

The following combinatorial lemma connects the property of being
\kl{well-quasi-ordered} to a property of the \kl{periodic lengths} of words in
a language, based on the assumption that some factors can be iterated.

@@

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the values of $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that $w[\vec{n_i}]$ is actually an \kl{infix} of $u_{\ell}
    v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    Even more,
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    \end{itemize}
    In all those cases, we conclude using \cref{powers-infixes:cor}
    that there exists $x,y \in \Sigma^+$ of size at most $s$, and 
    a number $1 \leq t \leq k$ such that
    $v_i^{n_i} \in \InfPeriodChain{x}$ for all $1 \leq i \leq t$,
    and
    $v_i^{n_i} \in \InfPeriodChain{y}$ for all $t < i \leq k$.
    In particular,
    $w[\vec{n_i}] \in \InfPeriodChain{x} u_{t} \InfPeriodChain{y}$.

@@

We are now ready to restate and prove our main theorem.

@@

\begin{corollary}
    Given a regular language $L$, it is decidable whether
    $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}
\begin{proof}
    It suffices to decide whether the equation 
    \cref{infix-automata:eq} holds for the language $L$.
    This is an inclusion of regular languages, which is decidable.
\end{proof}

@@

\AP Let us recall that over an alphabet $(\Sigma, =)$ a \kl{subword embedding}
between two words $u \in \Sigma^*$ and $v \in \Sigma^*$ is a function $\rho
\colon \range{\card{u}} \to \range{\card{v}}$ such that $u_i = v_{\rho(i)}$ for
all $i \in \range{\card{u}}$. We write $\intro*\HigEmb(u,v)$ the set of all
\kl{subword embeddings} between $u$ and $v$. It may be useful to notice that
the set of finite words over $\Sigma$ forms a category when we consider
\kl{subword embeddings} as morphisms. Given a \kl{subword embedding} $f \colon
u \to v$ between two words $u$ and $v$, there exists a unique decomposition $v
= \GapWord{f}{0} u_1 \GapWord{f}{1} \cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$
where $\GapWord{f}{i} = v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$,
$\GapWord{f}{k} = v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$.
We say that $\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$.

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings},
    and $\canrun \colon R \to (\Sigma \cup \set{\varepsilon})^*$ is a 
    function computing a \emph{canonical decompostion} of a run.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, we define
    the \intro{gap language} of $r$ at position $i$ as $\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item $(R, E)$ forms a category.
        \item $(R, \leq_E)$ is a well-quasi-ordered set.
        \item For all $r, s, t \in R$,
            for all $f \in E(r,s)$,
            for all $g \in E(r,t)$,
            and for all $0 \leq i_0 \leq \card{\canrun(r)}$,
            there exists a run $z \in R$ and morphisms
            $h \in E(s,z)$ and $k \in E(t,z)$
            satisfying
            $h \circ f = k \circ g$,
            $\GapLanguage{h \circ f}{i} = \GapLanguage{f}{i} \GapLanguage{g}{i}$
            or 
            $\GapLanguage{h \circ f}{i} = \GapLanguage{g}{i} \GapLanguage{f}{i}$
            for every $0 \leq j \leq n$,
            and
            $\GapLanguage{h \circ f}{i_0} = \GapLanguage{f}{i_0} \GapLanguage{g}{i_0}$.
    \end{enumerate}

    The \intro{amalgamation language} of such a system
    is the set of all words $w \in \Sigma^*$ such that
    there exists a run $r \in R$
    such that the concatenation of the letters (and possibly the empty word) in
    $\canrun(r)$ equals $w$.
\end{definition}

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing" of runs together to obtain new runs.

Let us recall some examples of languages that can be recognized by
\kl{amalgamation systems} given by the authors of the original paper: regular
languages \cite[Theorem 5.3]{ASZZ24}, VASS as a consequence of \cite[Theorem
5.5]{ASZZ24}, context-free languages as a consequence of \cite[Theorem
5.10]{ASZZ24}.

@@

    Let us consider the set $M$ of minimal runs for the relation $\leq_E$,
    which is finite because the latter is a \kl{well-quasi-ordering}. By
    definition, for every word $w \in L$, there exists $r \in M$, $s \in R$,
    and $f \in E(r,s)$ such that $w = \canrun(s)$.
    In particular, we conclude that
    \begin{equation*}
        L \subseteq \bigcup_{r \in M} 
        \left(
        \GapLanguage{r}{0}
        \prod_{i = 1}^{\card{\canrun(r)}-1} \canrun(r)_i \GapLanguage{r}{i} 
        \right)
        \quad .
    \end{equation*}

    Let $u$ and $v$ be two words in the \kl{gap language}
    $\GapLanguage{r}{\ell}$ for some $r \in M$ and $0 \leq \ell \leq
    \card{\canrun(r)}$. We have corresponding runs $s_1, s_2 \in R$ and
    embeddings $f_1 \in E(r,s_1)$ and $f_2 \in E(r,s_2)$ such that
    $\GapWord{f_1}{\ell} = u$ and $\GapWord{f_2}{\ell} = v$. Using the
    amalgamation property, we can inductively construct runs $r_k \in R$ and
    embeddings $f_k \in E(r,s_k)$ such that $\GapWord{f_k}{\ell} = u v^k u$,
    and $\GapWord{f_k}{i}$ is obtained by concatenating in some order two
    occurrences of $\GapWord{f_1}{i}$ and $k$ occurrences of
    $\GapWord{f_2}{i}$. Because $L$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}, there exists $i < j$ such that $\yieldrun(r_i) \infleq
    \yieldrun(r_j)$. Assuming that $\card{\canrun(r)}$ is greater than $2$, and
    that we have extracted an increasing sequence of indices so that
    $\alpha(i+1) > \factorial{\card{\yieldrun(r_i)}}$, we conclude that $u
    v^{\alpha(i)} u$ is an \kl{infix} of $\yieldrun(r_j)$.
    Because of the number of repetitions, a \emph{large} infix of $v^{\alpha(i)}$ must be
    an infix of some $\GapWord{f_2}{p}^{\alpha(j)}$.
    We conclude that $\Gapword{f_2}{p}^{\alpha(j)}$ and $v$
    have a common period. Refining the analysis, we also conclude
    that $uv^{q}$ and $\GapWord{f_2}{p}^{\alpha(j)}$ have the same period,
    which proves that $uv^{q}u$ belongs to some $\InfPeriodChain{x}$
    \textbf{todo}.

>>>>

\AP
In this section, we study languages equipped with the \kl{infix relation}. As
opposed to the \kl{prefix} and \kl{suffix} relations, the \kl{infix relation}
can yield to very complicated \kl{well-quasi-ordered} languages. Formally, the
upcoming \cref{infix-embedding:thm} shows that \emph{any} countable
partial-ordering with finite initial segments can be embedded into the infix
relation of a language. To make the former statement precise,
Let us recall that an \intro{order embedding} from a quasi-ordered set $(X,
\preceq)$ into a quasi-ordered set $(Y, \preceq')$ is a function $f \colon X
\to Y$ such that for all $x, y \in X$, $x \preceq y$ if and only if $f(x)
\preceq' f(y)$. When such an embedding exists, we say that $X$ \reintro{embeds
into} $Y$. We say that a quasi-ordered set $(X, \preceq)$ is a \intro{partial
ordering} whenever the relation $\preceq$ is antisymmetric, that is $x \preceq
y$ and $y \preceq x$ implies $x = y$.

@@

    Conversely, let us prove that \cref{infix-embedding-count:item} implies
    \cref{infix-embedding-embeds:item}. To that end, let us first define
    inductively a sequence $\seqof{X_n}$ of subsets of $X$ as follows: $X_0$ is
    the set of minimal elements of $X$, and $X_{n+1}$ is the set of minimal
    elements of $X \setminus \bigcup_{i \leq n} X_i$. Because initial segments
    of $X$ are finite, we know that $X = \bigcup_{n \in \Nat} X_n$. To
    simplify notations, we will also use $Y_n \defined \bigcup_{i \leq n} X_i$.
    We also let $L_0 \defined a b^* a$, and 
    $L_{n+1} \defined L_n \cup c^{n+1} b^* c^{n+1} (L_n c^{n+1})^*$.
    Now, let us inductively define \kl{embeddings} $f_n \colon (Y_n,\preceq) \to
    (L_n, \infleq)$.

@@

\begin{theorem}[restate=infix-finite-automata:thm,label=infix-finite-automata:thm]
    Let $L \subseteq \Sigma^*$ be a language recognized by a finite automaton.
    Then $L$ is \kl{well-quasi-ordered} by the \kl{infix relation} if and only if $L$ is
    a finite union of \kl{chains} for the \kl{infix relation}.
\end{theorem}

@@

The upcoming \cref{powers-infixes:cor} is based on the observation that if $S$
is a \kl{chain} for the \kl{suffix relation} and $P$ is a \kl{chain} for the
\kl{prefix relation}, then $SP$ is a \kl{chain} for the \kl{infix relation}.

@@

The following combinatorial lemma connects the property of being
\kl{well-quasi-ordered} to a property of the \kl{periodic lengths} of words in
a language, based on the assumption that some factors can be iterated. It is
the core result that powers the analysis done in the upcoming
\cref{infix-finite-automata:thm,infix-amalgamation:thm}.

@@

    Let us assume that $k \geq 2$ in the following proof for symmetry purposes,
    and argue later on that when $k = 1$ the same argument goes through.
    Because $L$ is \kl{well-quasi-ordered} by the \kl{infix relation}, there
    exists $i < j$ such that $w[\vec{n_i}]$ is an \kl{infix} of $w[\vec{n_j}]$.
    Now, because of the chosen values for $\vec{n_j}$, there exists $1 \leq \ell \leq
    k-1$ such that $w[\vec{n_i}]$ is actually an \kl{infix} of $u_{\ell}
    v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    Even more,
    one of the three following equations holds:
    \begin{itemize}
        \item $w[\vec{n_i}] \infleq v_{\ell}^{n_{j,\ell}} u_{\ell+1} v_{\ell+1}^{n_{j,\ell+1}}$,
        \item $w[\vec{n_i}] \infleq u_{\ell}
            v_{\ell}^{n_{j,\ell}}$,
        \item $w[\vec{n_i}] \infleq
            v_{\ell+1}^{n_{j,\ell+1}} u_{\ell+2}$.
    \end{itemize}
    In all those cases, we conclude using \cref{powers-infixes:cor}
    that there exists $x,y \in \Sigma^+$ of size at most $s$, and 
    a number $1 \leq t \leq k$ such that
    $v_i^{n_i} \in \InfPeriodChain{x}$ for all $1 \leq i \leq t$,
    and
    $v_i^{n_i} \in \InfPeriodChain{y}$ for all $t < i \leq k$.
    In particular,
    $w[\vec{n_i}] \in \InfPeriodChain{x} u_{t} \InfPeriodChain{y}$.

@@

We are now ready to restate and prove our main theorem, proven by combining
\cref{pumping-periods:lem} together with classical pumping arguments on finite
state automata.

@@

\begin{corollary}
    Given a regular language $L$, it is decidable whether
    $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{corollary}
\begin{proof}
    It suffices to decide whether 
    \cref{infix-automata:eq} holds for the language $L$.
    This is an inclusion of regular languages, which is decidable.
\end{proof}

Let $(X,\preceq)$ be a quasi-ordered set, and $L \subseteq X$ be such that $(L,
\preceq)$ is \kl{well-quasi-ordered}. It is not true in general that
$(\dwset{L}, \preceq)$ is \kl{well-quasi-ordered}. In the case of $(\Sigma^*,
\infleq)$ a typical example is to start from an infinite \kl{antichain} $A$,
together with an enumeration $\seqof{w_i}$ of $A$, and build the language $L
\defined \setof{ \prod_{i = 0}^n w_i }{ i \in \Nat }$. By definition, $L$ is a
\kl{chain} for the \kl{infix} ordering, hence \kl{well-quasi-ordered}. However,
$\dwset[\infleq]{L}$ contains $A$, and is therefore not
\kl{well-quasi-ordered}. As a fun corollary of our analysis, we conclude that
this particular example cannot be realized by a regular language $A$.

\begin{corollary}
    Given a regular language $L$, it is \kl{well-quasi-ordered} 
    for the \kl{infix relation} if and only if 
    $\dwset[\infleq] L$ is.
\end{corollary}

@@

\AP Let us recall that over an alphabet $(\Sigma, =)$ a \kl{subword embedding}
between two words $u \in \Sigma^*$ and $v \in \Sigma^*$ is a function $\rho
\colon \range{\card{u}} \to \range{\card{v}}$ such that $u_i = v_{\rho(i)}$ for
all $i \in \range{\card{u}}$. We write $\intro*\HigEmb(u,v)$ the set of all
\kl{subword embeddings} between $u$ and $v$. It may be useful to notice that
the set of finite words over $\Sigma$ forms a category when we consider
\kl{subword embeddings} as morphisms, which is a fancy way to state that
$\mathrm{id} \in \HigEmb(u,u)$ and that $f \circ g \in \HigEmb(u,w)$ whenever
$g \in \HigEmb(u,v)$ and $f \in \HigEmb(v,w)$, for any choice of words
$u,v,w \in \Sigma^*$.

\AP Given a \kl{subword embedding} $f \colon u \to v$ between two words $u$ and
$v$, there exists a unique decomposition $v = \GapWord{f}{0} u_1 \GapWord{f}{1}
\cdots \GapWord{f}{k-1} u_k \GapWord{f}{k}$ where $\GapWord{f}{i} =
v[f(i)+1:f(i+1)-1]$ for all $1 \leq i \leq k-1$, $\GapWord{f}{k} =
v[f(k)+1:\card{v}]$, and $\GapWord{f}{0}   = v[1: f(1)-1]$. We say that
$\intro*\GapWord{f}{i}$ is the $i$-th \intro{gap word} of $f$. These \kl{gap
words} will be useful to describe how and where runs of a system (described by
words) can be combined.

\begin{definition}
    An \intro{amalgamation system}
    is a triple $(\Sigma, R, E, \canrun)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    and 
    $E$ is a set of so-called \emph{run embeddings},
    and $\canrun \colon R \to (\Sigma \uplus \set{\varepsilon})^*$ is a 
    function computing a \emph{canonical decomposition} of a run.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{Word Embeddings as Morphisms.} For all $r, s \in R$,
            $E(r,s) \subseteq \HigEmb(\canrun(r), \canrun(s))$.
        \item \emph{$(R, E)$ Forms a Category.}
            For all $r,s,t \in R$,
            $\mathrm{id} \in E(r,r)$,
            and whenever $f \in E(r,s)$ and $g \in E(s,t)$,
            then $g \circ f \in E(r,t)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \leq_E)$ is a well-quasi-ordered set,
            where $r \leq_E s$ if and only if $E(r,s) \neq \emptyset$.
        \item \emph{Controlled Amalgamation.}
            For all $r, s, t \in R$,
            for all $f \in E(r,s)$,
            for all $g \in E(r,t)$,
            and for all $0 \leq i_0 \leq \card{\canrun(r)}$,
            there exists a run $z \in R$ and morphisms
            $h \in E(s,z)$ and $k \in E(t,z)$
            satisfying
            $h \circ f = k \circ g$,
            $\GapLanguage{h \circ f}{i} = \GapLanguage{f}{i} \GapLanguage{g}{i}$
            or 
            $\GapLanguage{h \circ f}{i} = \GapLanguage{g}{i} \GapLanguage{f}{i}$
            for every $0 \leq j \leq n$,
            and
            $\GapLanguage{h \circ f}{i_0} = \GapLanguage{f}{i_0} \GapLanguage{g}{i_0}$.
    \end{enumerate}

    The \intro{amalgamation language} of such a system
    is the set of all words $w \in \Sigma^*$ such that
    there exists a run $r \in R$
    such that the concatenation of the letters of its
    canonical decomposition, written $\intro*\yieldrun(r)$,
    equals $w$.
\end{definition}

Intuitively, the definition of an amalgamation system allows the comparison of
runs, and the proper ``gluing" of runs together to obtain new runs. Let us
recall some examples of languages that can be recognized by \kl{amalgamation
systems} given by the authors of the original paper: regular languages
\cite[Theorem 5.3]{ASZZ24}, VASS as a consequence of \cite[Theorem
5.5]{ASZZ24}, context-free languages as a consequence of \cite[Theorem
5.10]{ASZZ24}. For this paper to be self-contained, we will also recall how
runs of a finite state automaton can be understood as an \kl{amalgamation
system}.

\begin{example}
    Let $A = (Q, \delta, q_0, F)$ be a finite state automaton over a finite
    alphabet $\Sigma$. Let $\Delta$ be the set of transitions $(q_1, a, q_2)
    \in Q \times \Sigma \times Q$,
    and $R \subseteq \Delta^*$ be the set of 
    words over transitions that start with the initial state $q_0$,
    end in a final state $q_f \in F$, and such that the end state of a
    letter is the start state of the following one.

    The canonical decomposition $\canrun \colon R \to \Sigma^*$
    is defined as a morphism from $\Delta^*$ to $\Sigma^*$
    that maps $(q,a,p)$ to $a$.

    Finally, given two runs $r$ and $s$ of the automaton,
    let $E(r,s) = \HigEmb(\canrun(r), \canrun(s))$, that is, we do not
    remove morphisms.

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    TODO.
\end{proof}

@@

    Let us consider the set $M$ of minimal runs for the relation $\leq_E$,
    which is finite because the latter is a \kl{well-quasi-ordering}. By
    definition, for every word $w \in L$, there exists $r \in M$, $s \in R$,
    and $f \in E(r,s)$ such that $w = \canrun(s)$.
    In particular, we conclude that
    \begin{equation*}
        L \subseteq \bigcup_{r \in M} 
        \yieldrun \left(
        \GapLanguage{r}{0}
        \prod_{i = 1}^{\card{\canrun(r)}-1} \canrun(r)_i \GapLanguage{r}{i} 
        \right)
        \quad .
    \end{equation*}

    Let $u$ and $v$ be two words in the \kl{gap language}
    $\GapLanguage{r}{\ell}$ for some $r \in M$ and $0 \leq \ell \leq
    \card{\canrun(r)}$. We have corresponding runs $s_1, s_2 \in R$ and
    embeddings $f_1 \in E(r,s_1)$ and $f_2 \in E(r,s_2)$ such that
    $\GapWord{f_1}{\ell} = u$ and $\GapWord{f_2}{\ell} = v$. Using the
    amalgamation property, we can inductively construct runs $r_k \in R$ and
    embeddings $f_k \in E(r,s_k)$ such that $\GapWord{f_k}{\ell} = u v^k u$,
    and $\GapWord{f_k}{i}$ is obtained by concatenating in some order two
    occurrences of $\GapWord{f_1}{i}$ and $k$ occurrences of
    $\GapWord{f_2}{i}$. Because $L$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}, there exists $i < j$ such that $\yieldrun(r_i) \infleq
    \yieldrun(r_j)$. Assuming that $\card{\canrun(r)}$ is greater than $2$, and
    that we have extracted an increasing sequence of indices so that
    $\alpha(i+1) > \factorial{\card{\yieldrun(r_i)}}$, we conclude that $u
    v^{\alpha(i)} u$ is an \kl{infix} of $\yieldrun(r_j)$.
    Because of the number of repetitions, a \emph{large} infix of $v^{\alpha(i)}$ must be
    an infix of some $(\GapWord{f_2}{p})^{\alpha(j)}$.
    We conclude that $(\GapWord{f_2}{p})^{\alpha(j)}$ and $v$
    have a common period. Refining the analysis, we also conclude
    that $uv^{q}$ and $(\GapWord{f_2}{p})^{\alpha(j)}$ have the same period,
    which proves that $uv^{q}u$ belongs to some $\InfPeriodChain{x}$
    \textbf{todo}.

### 2024-08-15 src/introduction.tex
typos

A \intro{well-quasi-ordered} set is a set $X$ equipped with a quasi-order
$\preceq$ such that every infinite sequence $\seqof{x_n}$ of elements taken in
$X$ contains a pair $x_i \preceq x_j$ with $i < j$. Well-quasi-orderings serve
as a core combinatorial tool powering many termination arguments, and was
successfully applied to the verification of infinite state transition systems
\cite{ABDU96,ABDU98}. One of the appealing properties of well-quasi-orderings
is that they are closed under many operations, such as taking products, finite
unions, and finite powersets \cite{SCSC12}. More surprisingly, the class of
well-quasi-ordered sets is also stable under the operation of taking finite
words and finite trees labelled by elements of a well-quasi-ordered set
\cite{HIG52,KRU72}.

@@

\AP
However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. In our, case we
consider alphabets $\Sigma$ that are finite and equipped with the equality
relation. In this setting, the three alternatives we consider are the
\intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with $uw
= v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there exists
$w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---.
Unfortunately, it is an easy check that all of these constructions are not
well-quasi-ordered as soon as the alphabet contains two non-equivalent letters.
This is not the end of the road, because some \emph{subsets} of $\Sigma^*$
might still be \kl{well-quasi-ordered} by these relations, the simplest example
being the case of finite sets of (finite) words which are all
\kl{well-quasi-ordered} regardless of the ordering considered. This raises the
question of characterizing the subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} with respect to the \kl{prefix relation} (respectively,
the \kl{suffix relation} or the \kl{infix relation}), and providing suitable
decision procedures.

\subparagraph{Contribution} 
\AP
In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain an infinite \kl{antichain}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$.

@@

In the case of the \kl{infix relation}, the situation is more complex, and we
show that \emph{any} countable quasi-ordering with finite initial segments can
be embedded into the \kl{infix relation} of a language over a binary alphabet
$\Sigma \defined \set{a,b}$. As an example, on can embed $(\Sigma^*, \higleq)$
into $(\Sigma^*, \infleq)$ where $\higleq$ is the \kl{subword relation} and
$\infleq$ is the \kl{infix relation}. This strongly contrasts with the
\kl{prefix relation}, and suggests that a meaningful characterization of
\kl{well-quasi-ordered} languages under the \kl{infix relation} is unlikely to
be found.

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}, a recently introduced notion that abstracts the
notion of systems whose runs can be suitably ``pumped'' \cite{ASZZ24}. Under
this relatively mild assumption, we also prove that \kl{well-quasi-ordered}
languages under the \kl{infix relation} are precisely the finite unions of
\kl{chains}.

\subparagraph{Related work} The study of alternative \kl{well-quasi-ordered}
relations over finite words is far from new. For instance, orders obtained by
so-called \emph{derivation relations} where already analysed by Bucher,
Ehrenfeucht, and Haussler \cite{BUEUD85}. These results were later extended by
D'Alessandro and Varricchio \cite{ALVA03,ALVA06}. However, in all those cases
the orderings are \emph{multiplicative}, that is, if $u_1 \preceq v_1$ and $u_2
\preceq v_2$ then $u_1u_2 \preceq v_1v_2$. This assumption does not hold for
the \kl{prefix}, \kl{suffix}, and \kl{infix} relations.

@@

\subparagraph{Acknowledgements} We would like to thank participants of the 2024
edition of Autobóz for their helpful comments and discussions.

>>>>

A \intro{well-quasi-ordered} set is a set $X$ equipped with a quasi-order
$\preceq$ such that every infinite sequence $\seqof{x_n}$ of elements taken in
$X$ contains a pair $x_i \preceq x_j$ with $i < j$. Well-quasi-orderings serve
as a core combinatorial tool powering many termination arguments, and was
successfully applied to the verification of infinite state transition systems
\cite{ABDU96,ABDU98}. One of the appealing properties of well-quasi-orderings
is that they are closed under many operations, such as taking products, finite
unions, and finite powersets \cite{SCSC12}. Perhaps more surprisingly, the
class of well-quasi-ordered sets is also stable under the operation of taking
finite words and finite trees labelled by elements of a well-quasi-ordered set
\cite{HIG52,KRU72}.

@@

\AP However, there are many other natural orderings on words that could be
considered in the context of \kl{well-quasi-orderings}. Let us restrict
ourselves to a simplified case where the alphabet is finite and equipped with
the equality relation. In this setting, the three alternatives we consider are
the \intro{prefix relation} --- $u \intro*\prefleq v$ if there exists $w$ with
$uw = v$ ---, the \intro{suffix relation} --- $u \intro*\suffleq v$ if there
exists $w$ such that $wu = v$ ---, and the \intro{infix relation} --- $u
\intro*\infleq v$ if there exists $w_1,w_2$ such that $w_1 u w_2 = v$ ---. Note
that these three relations straightforwardly generalize to infinite
quasi-ordered alphabets. Unfortunately, it is an easy check that all of these
constructions are not well-quasi-ordered as soon as the alphabet contains two
non-equivalent letters. This is not the end of the road, because some
\emph{subsets} of $\Sigma^*$ might still be \kl{well-quasi-ordered} by these
relations, the simplest example being the case of finite sets of (finite) words
which are all \kl{well-quasi-ordered} regardless of the ordering considered.
This raises the question of characterizing the subsets $L \subseteq \Sigma^*$
that are \kl{well-quasi-ordered} with respect to the \kl{prefix relation}
(respectively, the \kl{suffix relation} or the \kl{infix relation}), and
providing suitable decision procedures.

\subparagraph{Contribution} 
\AP
In this paper, we focus on words over a finite
alphabet $\Sigma$, and characterize subsets $L \subseteq \Sigma^*$ that are
\kl{well-quasi-ordered} by the \kl{prefix relation}, the \kl{suffix relation},
and the \kl{infix relation}. Let us remark that these word orderings are all
\intro{well-founded}, i.e., they do not contain infinite \intro{decreasing
sequences} $\seqof{x_n}$ such that $x_{n+1} \prec x_n$ for all $n \in \Nat$.
Let us recall that the key difference between \kl{well-quasi-orderings} and the
maybe more usual notion of well-founded orderings is that the latter allows the
existence of infinite \intro{antichains}, that is, sets of pairwise
incomparable elements. In fact, a set $X$ is \kl{well-quasi-ordered} if and
only if it is \kl{well-founded} and does not contain infinite \kl{antichains}
\cite{KRU72}.  As a consequence, our study focuses on detecting infinite
\kl{antichains} in languages $L \subseteq \Sigma^*$.

@@

In the case of the \kl{infix relation}, the situation is more complex, and we
show that \emph{any} countable quasi-ordering with finite initial segments can
be \kl{embedded} into the \kl{infix relation} of a language over a binary
alphabet $\Sigma \defined \set{a,b}$. As an example, on can identify
$(\Sigma^*, \higleq)$ with a subset of $(\Sigma^*, \infleq)$ where $\higleq$ is
the \kl{subword relation} and $\infleq$ is the \kl{infix relation}. This
strongly contrasts with the \kl{prefix relation}, and suggests that a
meaningful characterization of \kl{well-quasi-ordered} languages under the
\kl{infix relation} is unlikely to be found.

We circumvent this issue by adding some regularity assumptions on the languages
$L \subseteq \Sigma^*$ that we consider. This can be done by assuming that $L$
is recognized by a finite automaton, or more generally that $L$ is the language
of an \kl{amalgamation system}, a recently introduced notion that abstracts the
notion of systems whose runs can be suitably ``pumped'' \cite{ASZZ24}. As
examples \kl{amalgamation systems} covers the case of regular languages,
context free languages, or even languages recognized by ``vector addition
systems with state" \cite{ASZZ24}. Under this relatively mild assumption, we
also prove that \kl{well-quasi-ordered} languages under the \kl{infix relation}
are precisely the finite unions of \kl{chains}.

\subparagraph{Related work} The study of alternative \kl{well-quasi-ordered}
relations over finite words is far from new. For instance, orders obtained by
so-called \emph{derivation relations} where already analysed by Bucher,
Ehrenfeucht, and Haussler \cite{BUEUD85}, and were later extended by
D'Alessandro and Varricchio \cite{ALVA03,ALVA06}. However, in all those cases
the orderings are \emph{multiplicative}, that is, if $u_1 \preceq v_1$ and $u_2
\preceq v_2$ then $u_1u_2 \preceq v_1v_2$. This assumption does not hold for
the \kl{prefix}, \kl{suffix}, and \kl{infix} relations.

@@

\subparagraph{Acknowledgements} We would like to thank participants of the 2024
edition of \kl{Autobóz} for their helpful comments and discussions.
